# Reproducible Research: Peer Assessment 1  
#### Personal Activity Data Analysis Report  

=========================================================================================

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices. But this data remains underutilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This analysis makes use of data from a personal activity monitoring device. This device collects data at 5-minute intervals throughout the day. The data consists of two-month worth of data collected during October and November, 2012. The data includes the number of steps taken in 5-minute intervals each day.

## Data
The raw dataset can be obtained through this link [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) as a zipped file. However, the code presented in this report downloads the dataset automatically for you to reporduce the research.

The raw dataset is comprised of the following variables:

* **steps**: Number of steps taken in a 5-minute interval (missing values are coded as NA).
* **date**: The date on which the measurement was taken in YYYY-MM-DD format.
* **interval**: Identifier for the 5-minute interval in which measurement was taken.

The dataset is stored in a comma-separated-value (CSV) file and there are a total of **17,568** observations in this dataset.

## Loading and preprocessing the data
#### 1. Load the data

The analysis starts by downloading the zipped raw data file. Then the file is unzipped.

```{r Download and unzip data, echo=TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
dataFileZIP <- "./repdata_data_activity.zip"

if (file.exists(dataFileZIP) == FALSE) {
        download.file(fileURL, destfile = dataFileZIP)
}
unzip(dataFileZIP)
```

Load the raw data file. (The first 10 rows of the raw data is shown below).

```{r Load raw data,echo=TRUE}
RawData<-read.csv("./activity.csv")
head(RawData,10)
```

#### 2. Processing/transforming the data is not necessary at this stage.

#### Loading necessary R Packages
The analysis uses some packages that need to be installed before running the code. The code below loads these packages. Yet, you need to make sure they are installed in your R version before loading.

```{r Loading packages,echo=TRUE}
library(dplyr)
library(ggplot2)
```

## What is mean total number of steps taken per day?
For this part of the analysis, the missing values in the dataset are ignored.

#### 1. Calculate the total number of steps taken per day
The code uses the 'dplyr' package to calculate the total number of steps taken per day.
(The first 5 rows of the resulted dataset is shown below).

```{r Steps per day,echo=TRUE}
df0<-group_by(RawData,date)
StepsPerDay<-summarize(df0,DaySteps=sum(steps,na.rm = TRUE))
head(StepsPerDay,5)
```

#### 2. Make a histogram of the total number of steps taken each day.
The histogram is generated using the ggplot2 plotting system.

```{r Histogram,echo=TRUE}
g<-ggplot(StepsPerDay,aes(x=DaySteps))
g+geom_histogram()+
        labs(x="Steps per Day",y="Frequency")+        
        theme(plot.title = element_text(size = 14, face = "bold", colour = "black", vjust = +1))+        
        ggtitle(expression(atop("Frequecy of total number of steps taken each day",
                                atop(italic("without imputing missing values")))))
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day.

```{r Mean and Median,echo=TRUE}
MeanSteps<-as.integer(mean(StepsPerDay$DaySteps,na.rm=TRUE))
MedianSteps<-as.integer(median(StepsPerDay$DaySteps,na.rm = TRUE))
```

The mean value of the total number of steps taken per day is **`r MeanSteps`**. Whereas the median value of the total number of steps taken per day is **`r MedianSteps`**.

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

In order to produce this plot we first summarize the raw data by creating the average steps per interval across all days. (The first 5 rows of the averaged data is shown below).

Then, we plot intervals versus average number of steps in a time-series plot.

```{r Average steps per interval,echo=TRUE}
df1<-group_by(RawData,interval)
AvgStepsPerInt<-summarize(df1,AvgSteps=mean(steps,na.rm = TRUE))

head(AvgStepsPerInt,5)

g<-ggplot(AvgStepsPerInt,aes(interval,AvgSteps))
g+geom_line()+
        labs(x="5-minute Interval",y="Avergae Number of Steps")+        
        theme(plot.title = element_text(size = 14, face = "bold", colour = "black", vjust = +1))+        
        ggtitle(expression(atop("Interval vs. Number of Steps",
                                atop(italic("Averaged across all days")))))
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r Interval of the max number of steps,echo=TRUE}
MxSteps<-max(AvgStepsPerInt$AvgSteps)
MxStepsInt<-as.integer(AvgStepsPerInt$interval[which(AvgStepsPerInt$AvgSteps==MxSteps)])
MxSteps<-as.integer(MxSteps)
```

The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is interval **`r MxStepsInt`** with a maximum number of steps of **`r MxSteps`** steps.

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r Missing values,echo=TRUE}
NAs<-sum(is.na(RawData$steps))
```

The total number of missing values in the dataset is **`r NAs`**.

#### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

We need to devise a strategy for filling in all of the missing values in the dataset.
The mean of steps taken across all days for the 5-minute interval corresponding to the missing value will be used as a strategy to fill in the missing values.

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

We first create a new dataset that is equal to the original dataset.

```{r New dataset,echo=TRUE}
ProData<-RawData
```

Then, we replace the NAs by their average 'steps' from the AvgStepsPerInt dataset for the same interval. (The first 5 rows of the new data set with filled NAs is shown below).

```{r Replacing NAs,echo=TRUE}
ProData$steps[ProData$interval==AvgStepsPerInt$interval & is.na(ProData$steps)]<-AvgStepsPerInt$AvgSteps

head(ProData,5)
```

#### 4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the analysis? What is the impact of imputing missing data on the estimates of the total daily number of steps?

For the new dataset (with NAs filled in), we will make a histogram of the total number of steps taken each day, then see the impact of imputing missing data on the estimates of the total daily number of steps.

First, we calculate the mean total number of steps taken per day. Then make the Histogram.

```{r Histogram with missing values filled,echo=TRUE}
df2<-group_by(ProData,date)
StepsPerDayPro<-summarize(df2,DaySteps=sum(steps,na.rm = TRUE))

head(StepsPerDayPro,5)

g<-ggplot(StepsPerDayPro,aes(x=DaySteps))
g+geom_histogram()+
        labs(x="Steps per Day",y="Frequency")+        
        theme(plot.title = element_text(size = 14, face = "bold", colour = "black", vjust = +1))+        
        ggtitle(expression(atop("Frequecy of total number of steps taken each day",
                                atop(italic("with imputing missing values")))))
```

Then, we calculate the mean and median of the total number of steps taken per day for the new dataset.

```{r Mean and Median of new dataset,echo=TRUE}
MeanStepsPro<-as.integer(mean(StepsPerDayPro$DaySteps,na.rm=TRUE))
MedianStepsPro<-as.integer(median(StepsPerDayPro$DaySteps,na.rm = TRUE))
```

The mean value of the total number of steps taken per day for the new dataset is **`r MeanStepsPro`**. And the median value is also **`r MedianStepsPro`**.

Notice that the mean and median for the imputed missing values data set are different from those in the first part (without imputing missing values). Both mean and median in the new dataset are higher than mean and median in the original dataset, respectively.

The impact of imputing missing data on the estimates of the total daily number of steps is that it eliminates skewness and bias in the data and converts the data distribution into unimodal probability distribution where mean and median are equal. Whereas the dataset with missing values exhibits skewness as implied by median being higher than the mean.

## Are there differences in activity patterns between weekdays and weekends?

Notice that we will use the dataset with the filled-in missing values for this part (the new dataset). The aim here is to see the differences in activity patterns between weekdays and weekends.

#### 1. create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
Here the assumption is that Saturday and Sunday are considered the weekend days.

First, we will create a variable in the new dataset named 'Day' presenting the day of the week. Then we will create the factor variable of 'weekday' and 'weekend' (DayType). (The first 5 rows of the new data set with Day Type is shown below)

```{r Weekday vs Weekend,echo=TRUE}
ProData$Day<- weekdays(as.Date(ProData$date),abbreviate = TRUE)
ProData$DayType<-as.factor(ifelse(ProData$Day %in% c("Sat","Sun"),"weekend","weekday"))

head(ProData,5)
```

#### 2. Make a panel plot containing a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

First, we need to summarize the new data by creating the average steps per interval per Day Type. Then, we plot intervals versus average number of steps into two panels; one for weekday and another for weekend.

```{r Plot weekday and weekend vs steps,echo=TRUE}
df3<-group_by(ProData,DayType,interval)
AvgStepsPerIntWd<-summarize(df3,AvgSteps=mean(steps))

head(AvgStepsPerIntWd,5)

g<-ggplot(AvgStepsPerIntWd,aes(interval,AvgSteps))
g+geom_line()+facet_grid(DayType~.)+
        labs(x="5-minute Interval",y="Avergae Number of Steps")+        
        theme(plot.title = element_text(size = 14, face = "bold", colour = "black", vjust = +1))+        
        ggtitle(expression(atop("Interval vs. Number of Steps",
                                atop(italic("Averaged across weekdays or weekends")))))
```

From the above plots we can conclude that activity patterns for weekdays and weekends are nearly similar. However, the below calculations of means can shed more light on the difference.

Let's calculate the mean of the average number of steps per interval for each of the weekdays and weekends.

```{r Weekday vs weekend statistics, echo=TRUE}
tapply(AvgStepsPerIntWd$AvgSteps,AvgStepsPerIntWd$DayType,mean)
```

Based on the results, the mean of number of steps per interval for weekdays is lower than that of the weekend. On average, this may imply that higher activities are performed during weekends than weekdays.

======================================END OF REPORT======================================